First of all, I would like to thank my supervisor, Dr. Xavier Serra, for giving me the opportunity to work in this fantastic environment, the Music Technology Group, and for his wise advises. Also, I want to give special thanks to Paco Gomez for teaching me how to be a researcher. This thesis does not have a specific co-supervisor, but along this journey I have met three great researchers and better persons who have helped me through my PhD and without whom this work would have not been possible, Mohamed Sordo, Vito Claudio Ostuni, and Oriol Nieto. 
Special thanks also to Frederic Font for all these years sharing office, research, music, and friendship.
A very important element of this thesis has been my collaboration with the TALN group. Everything started sharing pizza and water and now we have a lot of papers together and thousands of ideas. It was great collaborating and partying with Luis Espinosa-Anke and Francesco Barbieri. Also special thanks to Horacio Saggion for his wise advises, and Tommaso Di Noia, Aonghus Lawlor, and Michael O'Mahony for hosting me during my research stays at Politecnico de Bari and University College of Dublin. 
A special mention to Ichiro Fujinaga and Susan Weiss, who strongly believed in my research line. Also to my COFLA mates Emilia Gómez, Joaquin Mora, and José Miguel Díaz-Báñez, thanks for considering me a researcher from the first day.
To my Pandora managers Andreas Ehmann, Oscar Celma, and Fabien Gouyon, who trusted me and gave me the opportunity to apply my research in the real world.

I also want to thank other amazing MTG people who shared with me knowledge and laughs throughout these years. In no specific order, thanks to the SIC-refugio team
Sebastian Mealla, Álvaro Sarasúa, Panos Papiotis. To Rafael Caro, my musicologist mate. To Oriol Romaní and Juanjo Bosh for so much fun. To Alastair Porter and Dmitry Bogdanov for their experience and friendship. To Andrés Ferraro who shared my vision. To Gopala Koduri and our conversations about semantics. To Jordi Pons for initiating me in the deep learning cult. To my new roommates Eduardo Fonseca, Xavier Favory, and former roommates Gerard, Dara, Hector, Giuseppe, and Albin. To even more great people I met in the MTG, Cárthach, Ángel, Dani, Marius, Alfonso, Sergio, Zacharias, Olga, Perfe, Sergi, Rong, and Georgi. To former members, Sankalp, Sertan, Sercan, Ajay, Swapnil, Justin, Jordi, Julian, Julio, Nadine, and Martí. To Joan and Miguel from the TALN, Humberto from UCD, Manuel. To my Pandora mates Massimo, Theo, Chun, and Andreu. Sorry if I forgot to mention anyone.
Also thanks to Aurelio, Lydia, Sonia, Cristina, Magda, Jana, and Alba for helping me with the administrative work.
Special thanks to Obra Social "La Caixa" and their fellowship program, to believe in me and support this research.

Last but not least, this work would have been never been possible without the support of my wife Chiara, who followed me to Barcelona and helps me to pursue my dreams. Also to my lovely Olivia, I started my PhD and my paternity at the same time, and I am pretty sure these have been the happiest years of my life. Finally, many thanks to my parents, brothers and sisters, who helped me so much along my whole life, I am here thanks to them.

\vspace*{\fill}

\line(1,0){372}\\
\footnotesize
This thesis has been carried out at the Music Technology Group of Universitat Pompeu Fabra (UPF) in Barcelona, Spain, from October~2013 to September~2017, including a stay at Politecnico di Bari, Italy, from September~2014 to October~2014, at the Insight Centre for Data Analytics of University College of Dublin (UCD), Ireland, from May~2015 to July~2015, and at Pandora Media Inc., USA, from June~2016 to September~2016. It has been supervised by Dr. Xavier Serra.
Part of the work described in Chapter~\ref{sec:kb} has been conducted in collaboration with Luis Espinosa-Anke, both researchers in the Department of Information and Comunication Technologies of the Universitat Pompeu Fabra, Barcelona. Parts of the work described in Chapter~\ref{sec:graph-rec} has been conducted in collaboration with Vito Claudio Ostuni, during the research stay at Politecnico di Bari. 
Work in some parts of this thesis has also been carried out in collaboration with other researchers, including Mohamed Sordo, Oriol Nieto, Horacio Saggion, Tommaso Di Noia, Francisco Gómez, Frederic Font, Emilia Gómez, Aonghus Lawlor, and Joaquín Mora. 
The work in this thesis has been mainly supported by Obra Social "La Caixa" under their doctoral fellowship program, and by the Spanish Ministry of Economy and Competitiveness under the Maria de Maeztu Units of Excellence Programme (MDM-2015-0502). The research stay at UCD has been also funded by the Keystone COST Action IC1302 under grant number SFI/12/RC/2289. Parts of the work presented in this thesis were partly supported by the European Research Council under the European Union's Seventh Framework Program, as part of the CompMusic project (ERC grant agreement 267583), and by the COFLA2 research project (Proyectos de Excelencia de la Junta de Andalucia, FEDER P12-TIC-1362).
\normalsize